\hspace{1cm} Do syntactic and semantic similarity lead to interference effects? \newline Evidence from self-paced reading and event-related potentials using German

### 2024-08-11 paper/ResponseToReviewers.tex

\subsubsection*{Response to R1.2}
We have re-done the Figure to reflect our revised modeling. We hope that the scale of the new Figure \ref{fig:model_predictions_ERP} is easier to read (see below in our Response to R1.3).

\subsection*{Comment R1.3}
\begin{quote}
``In my reading, the discrepancy mainly comes from the fact that the decision to neglect syntactic interference is based on Bayes factors (see also below), but when matching the model predictions to actual results (p. 39-42), the match is done on posterior distributions, which do show a positive effect of interaction (the self-paced reading study) and an effect of syntactic interference (the self-paced reading study and the EEG study). \textbf{I think what the authors could do to at least partially address this issue is to be more explicit about model comparisons. They could leverage their use of Bayes factors to compare different computational models - the one proposed in this paper, which hard-codes that only the syntactic target is selected, the one based on Mertzen et al., possibly with varying priors, and, possibly, a null model which does not predict any interference (e.g., it hard-codes that only the target is retrieved).}''\end{quote}

\subsubsection*{Response to R1.3}
As suggested, we have implemented a null model (no interference), leaving us with a total of three computational models (the null model, the three-cues model by \textcite{mertzen} and our newly proposed two-cues model). We use Bayes factors to quantitatively compare these models. The result of this model comparison is that our new two-cues model outperforms the other two models. We have described the new modeling and model comparisons starting page \pageref{modeling}:

@@

\subsubsection*{Response to R2.2}
We agree with this statement and therefore refrained from interpreting the results at the critical region. To emphasize this point, we added the following to the manuscript on page \pageref{only_precritical}:\\

\begin{quote}
    \Paste{only_precritical}
\end{quote}

\subsection*{Comment R2.3}
\begin{quote}
``\textbf{(Furthermore, what's happening in regions 8-9? There is an increase in RTs in these regions in the high syntactic interference condition and a decrease in the low syntactic interference condition, while the trend for the remaining regions is quite similar. It's almost as if participants were able to discern the two syntactic conditions as early as region 8).}''
\end{quote}

\subsubsection*{Response to R2.3}
The participants were able to discern the two syntactic conditions at region 8 because this is the region were the conditions started to differ (`\dots erzählt hatte \dots' vs.\ `\dots von dem \dots'). We have added the words of an example item as x-axis labels to Figure \ref{fig:whole_sentence} to illustrate this clearly. For convenience, we paste the Figure here:

### 2024-08-12 paper/main.tex
Updated via the Open Science Framework

\label{modeling}\Copy{modeling}{\textcolor{blue}{The \citet{Lewis2005} model is generally used to predict reading times. Recently, \textcite{mertzen} presented simulations for the present design that predicted approximately equally sized effects of syntactic and semantic interference in the reading times of the critical verb. We found that the reading times at the critical verb were confounded by pre-critical effects, which were best explained by \textit{encoding} interference. Therefore, it makes no sense to compare the results from the present SPR experiment to predictions from the Lewis and Vasishth (2005) model of sentence processing as skilled memory \textit{retrieval}. Instead we use the Lewis and Vasishth (2005) model to generate predictions on the microvolt scale for ERP data. For this we adjust the scaling factor $F$ which usually links the activation of items in memory to retrieval times \citep{Lewis2005,anderson_etal_2004_ACT_R} to link memory activation to brain responses.} The modeling reported here is, to the best of our knowledge, the first time that the \citet{Lewis2005} model has been fit quantitatively to ERP data. We sampled the free scaling parameter $F$ from a \textcolor{blue}{truncated normal prior distribution ($Normal(0, 0.025)$) with a lower bound of 0 and an upper bound of 0.05. This prior reflects the a priori belief that effects will range from 0 to 5 $\mu V$ with higher likelihood for effect sizes below 1 $\mu V$ (see Figure \ref{fig:model_predictions_ERP}). This relatively uninformative prior was chosen because although there exists no previously published ERP data on the Van Dyke (2007) design it reflects the reasonable range of effect sizes in psycholinguistic ERP research.} After sampling the scaling parameter, simulated data was generated from the model while holding all other parameters in the model at fixed values. \textcolor{blue}{We decided to model N400 amplitude differences because Bayes factors provided stronger evidence for the semantic interference effect on the N400 than on the P600 in the present ERP study. To reflect that larger predicted effect sizes correspond to more negative N400 amplitudes, the predicted effects from the model} were multiplied with $-1$.

@@

It is informative to fit the model\textcolor{blue}{s} to the data and then examine the posterior predictive distribution of the \textcolor{blue}{ERPs} for each effect. 
We generated posterior predictive distributions for the effects of interest by fitting the $F$ parameter to the \textcolor{blue}{ERP} data. To avoid overfitting, we fitted $F$ only to the semantic interference effect (and not the syntactic interference effect or interaction). This was done by implementing an Approximate Bayesian Computation approach \parencite{SissonABC}: the model's predictions are generated iteratively, and only those parameter values for the scaling parameter $F$ are retained that predict values for the semantic interference effect that are close to the observed effect size $d$ in the \textcolor{blue}{ERP} data. `Close' is defined as $d \pm \,2\times SE$ ms; for details of this approach as applied to the cue-based retrieval model, see  \textcite{vasishth2020_abc}.

\textcolor{blue}{Out of 200,000 iterations of ABC algorithm, we obtained only 40,748 samples of $F$ for the 3-cues model, 28,718 for the 2-cues model and 6,615 for the 1-cue model. This is not surprising given
the relatively wide prior on $F$. In future work, the posterior distribution, with range [-0.5, -0.1]\,$\mu V$, could be used as a basis for redefining a tighter prior on the scaling parameter $F$. Figure \ref{fig:model_predictions_ERP} shows the  posterior predictive distributions
of the effects of interest in comparison to the prior predictive distribution and our ERP
data. The prior and posterior predictions from the model with three cues show equally large effects of syntactic and semantic interference and an interaction. In contrast, the model using two cues, including a syntactic cue that utilizes syntactic hierarchy, predicts no effect of syntactic interference and no interaction, but a semantic interference effect. The model using one cue predicts no interference effects. The prior and posterior predictions of all models overlap with the effects of interest in the data. }

\Copy{figpriorposteriordata}{
\begin{sidewaysfigure}[htpb]
    \caption{Prior and posterior predicted effect sizes on the microvolt scale, i.e., for negative ERPs, for syntactic and semantic interference and their interaction \textcolor{blue}{from three versions of the Lewis and Vasishth (2005) cue-based retrieval model.}
    %using the two cues \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. 
    The data with 95 \% credible intervals are shown in red.}\label{fig:model_predictions_ERP}
        \centering
         \includegraphics[width=\textwidth]{images/ERP_priorpred_vs_postpred_vs_data.png}
\end{sidewaysfigure}
}

\textcolor{blue}{To evaluate whether our new 2-cues model quantitatively outperforms the two other models in its predictive fit, we computed Bayes factors for the relevant model pairs under six different priors for the $F$ parameter 
($Normal$(0.01,0.01), 
$Normal$(0.01,0.02),
$Normal$(0.01,0.005),
$Normal$(0.02,0.01),
$Normal$(0.02,0.02),
$Normal$(0.02,0.005)). First, we computed the likelihood of the sampled $F$ parameters under each of the priors (LKL$_{F-prior}$) and the sampling prior (LKL$_{F-model}$). Then, the likelihood of the predicted semantic interference effect from the model was computed given the size of the observed semantic interference effect (LKL$_{predicted-effect}$). To obtain the marginal likelihood of the model, the product of LKL$_{predicted-effect}$ and LKL$_{F-prior}$ was divided by LKL$_{F-model}$. The marginal likelihood of the 2-cues model was then divided by the marginal likelihood of the three-cues and one-cue model, respectively. The resulting Bayes factors are shown in Figure \ref{fig:BF_compmods}. Compared to the null model with one cue, the Bayes factors provide extreme evidence for the new two-cues model. Therefore, we can conclude that the computational modeling shows that our data shows interference. Compared to the three-cues model proposed by \textcite{mertzen}, the Bayes factors provide moderate evidence for the new two-cues model. Therefore, the assumption that the relevant syntactic cue is the \{$\pm$subject-in-same-clause\} cue – rather than the previously assumed \{$\pm$grammatical subject\} cue – is moderately supported by these modeling results.
}

>>>>

\label{modeling}\Copy{modeling}{\textcolor{blue}{The \citet{Lewis2005} model is generally used to predict reading times. Recently, \textcite{mertzen} presented simulations for the present design that predicted approximately equally sized effects of syntactic and semantic interference in the reading times of the critical verb. We found that the reading times at the critical verb were confounded by pre-critical effects, which were best explained by \textit{encoding} interference. Therefore, it makes no sense to compare the results from the present SPR experiment to predictions from the Lewis and Vasishth (2005) model of sentence processing as skilled memory \textit{retrieval}. Instead we use the Lewis and Vasishth (2005) model to generate predictions on the microvolt scale for ERP data. For this we adjust the scaling factor $F$ which usually links the activation of items in memory to retrieval times \citep{Lewis2005,anderson_etal_2004_ACT_R} to link memory activation to brain responses.} The modeling reported here is, to the best of our knowledge, the first time that the \citet{Lewis2005} model has been fit quantitatively to ERP data. We sampled the free scaling parameter $F$ from a \textcolor{blue}{truncated normal prior distribution ($Normal_{lb=0,ub=0.05}(0.01, 0.01)$) with a lower bound of 0 and an upper bound of 0.05. This prior reflects the a priori belief that effect on the microvolt scale will range from 0 to 5 $\mu V$ with higher likelihood for effect sizes below 3 $\mu V$ (see Figure \ref{fig:model_predictions_ERP}). This relatively uninformative prior was chosen because although there exists no previously published ERP data on the Van Dyke (2007) design it reflects the reasonable range of effect sizes in psycholinguistic ERP research.} After sampling the scaling parameter, simulated data was generated from the model while holding all other parameters in the model at fixed values. \textcolor{blue}{We decided to model N400 amplitude differences because Bayes factors provided stronger evidence for the semantic interference effect on the N400 than on the P600 in the present ERP study. To reflect that larger predicted effect sizes correspond to more negative N400 amplitudes, the predicted effects from the model} were multiplied with $-1$.

@@

%It is informative to fit the model\textcolor{blue}{s} to the data and then examine the posterior predictive distribution of the \textcolor{blue}{ERPs} for each effect. We generated posterior predictive distributions for the effects of interest by fitting the $F$ parameter to the \textcolor{blue}{ERP} data. To avoid overfitting, we fitted $F$ only to the semantic interference effect (and not the syntactic interference effect or interaction). This was done by implementing an Approximate Bayesian Computation approach \parencite{SissonABC}: the model's predictions are generated iteratively, and only those parameter values for the scaling parameter $F$ are retained that predict values for the semantic interference effect that are close to the observed effect size $d$ in the \textcolor{blue}{ERP} data. `Close' is defined as $d \pm \,2\times SE$ ms; for details of this approach as applied to the cue-based retrieval model, see  \textcite{vasishth2020_abc}.

%\textcolor{blue}{Out of 200,000 iterations of ABC algorithm, we obtained only 40,748 samples of $F$ for the 3-cues model, 28,718 for the 2-cues model and 6,615 for the 1-cue model. This is not surprising given the relatively wide prior on $F$. In future work, the posterior distribution, with range [-0.5, -0.1]\,$\mu V$, could be used as a basis for redefining a tighter prior on the scaling parameter $F$. Figure \ref{fig:model_predictions_ERP} shows the  posterior predictive distributions of the effects of interest in comparison to the prior predictive distribution and our ERP data. The prior and posterior predictions from the model with three cues show equally large effects of syntactic and semantic interference and an interaction. In contrast, the model using two cues, including a syntactic cue that utilizes syntactic hierarchy, predicts no effect of syntactic interference and no interaction, but a semantic interference effect. The model using one cue predicts no interference effects. The prior and posterior predictions of all models overlap with the effects of interest in the data. }

\Copy{figpriorposteriordata}{
\begin{sidewaysfigure}[htpb]
    \caption{The prior predictions on the microvolt scale, i.e., for negative ERPs, for syntactic and semantic interference and their interaction \textcolor{blue}{from three versions of the Lewis and Vasishth (2005) cue-based retrieval model.}
    %using the two cues \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. 
    The data with 95 \% credible intervals are shown in red.}\label{fig:model_predictions_ERP}
        \centering
         \includegraphics[width=\textwidth]{images/ERP_priorpred_vs_postpred_vs_data.png}
\end{sidewaysfigure}
}

\textcolor{blue}{We evaluate whether our new two-cues model quantitatively outperforms the three-cues model and the one-cue model in its predictive performance on the semantic interference data. To evaluate the relative performance of the models, we use the Bayes factor analysis. We compute Bayes factors for the two-cues model compared with the three-cues model and the one cue model. The Bayes factor obtained for the two-cues model would imply the evidence in favor of this model when it is compared against the other two models.   
Since Bayes factors are sensitive to the choice of priors on the parameters of interest. We compute Bayes factors under six different priors for the latency factor parameter $F$,
$Normal$(0.01,0.01), 
$Normal$(0.01,0.02),
$Normal$(0.01,0.005),
$Normal$(0.02,0.01),
$Normal$(0.02,0.02), and 
$Normal$(0.02,0.005).

The procedure for computing Bayes factors is described as follows. The Bayes factor computation requires us to estimate the marginal likelihood of the competing models. We used the Monte Carlo integration method to estimate the marginal likelihood of each model. First, we chose an importance density for the parameter $F$ from which we draw $0.2$ million samples of $F$. Then, for each of these proposed parameter values, we compute the likelihood and prior density. The estimated marginal likelihood of a model will be given by

\begin{equation}
ML = \frac{1}{n} \sum_{i=1}{n} \mathcal{L}{F_i|y} p(F_i)  \text{ where } F_i \sim Normal_{lb=0,ub=0.05}(0.0.025) 
\end{equation}

\noindent where $y$ is the observed data, i.e., observed semantic interference effect in our case, $F_i$ is the $i^{th}$ proposal from the importance density $Normal_{lb=0,ub=0.05}(0.0.025) $. The term $\mathcal{L}{F_i|y}$ gives the likelihood of proposal $F_i$ given the data $y$ and $p(F_i)$ gives the prior density of $F_i$. We compute marginal likelihoods under different prior assumptions $p(F)$.

The LV05 model is a complex, non-deterministic model and it is difficult the express the model's likelihood anallytically without drastically simplifying the model cite[e.g., in ][]{NicenboimRetrieval2018,lisson2020computational}. In absence of an analytically-expressed likelihood function, we use Approximate Bayesian Computation (ABC)~\citep{sisson2018handbook,palestro2018likelihood} approach to estimate approximate likelihood for a given value of $F$. Under this approach, we compare the model generated data conditioned on $F=F_i$ and the actual observed data $y$. If the model-generated data $x_{i} \sim Model(F_i)$ for $F_i$ is close to the actual data $y$, then $F_i$ has higher likelihood and vice versa. The likelihood term can be rewritten as $\mathcal{L}{F_i|y} = \Psi(d(x_i, y), 0 , \delta)$, where $\Psi(.|\delta)$ represent a density kernel that weighs a proposal $F_i$ based on ditance between model-generated data and actual data, $d(x_i,y)$.

Using the above method, we estimate marginal likelihood for each candidate model under six different prior assumptions. The Bayes factor in favor of a model M1
compared to a model M2, i.e., BF12 is computed as the ratio
of the marginal likelihoods of M1 and M2. The Bayes factor BF12 represents the extent to which the model M1 is more
likely than M2 given the data. Following the convention from
Jeffreys (1939/1998), a Bayes factor value of larger than 10
is interpreted as strong evidence in favor of M1 and a value
between 3 and 10 is interpreted as moderate evidence in favor
of M1.

From the marginal likelihoods, we compute Bayes factors for the two-cues model compared against the three-cues model and the one-cue model. That is, the marginal likelihood of the 2-cues model $\mathrm{M1}$ was divided by the marginal likelihood of the three-cues model $\mathrm{M2}$ and one-cue model $\mathrm{M3}$ to obtain the Bayes factors $BF_{12}$ and $BF_{13}$ respectively. Following the convention from \citep{jeffreys1998theory}, if the Bayes factor value is larger than $10$. it would be interpreted as strong evidence in favor of the two-cues model and if the Bayes factor is between $3$ and $10$, it is interpreted as moderate evidence in favor of the two-cues model compared against the competing model.

The Bayes factors under different priors on $F$ are shown in Figure \ref{fig:BF_compmods}. Compared to the null model, the one-cue model, the Bayes factors indicate strong evidence for the new two-cues model. Therefore, we can conclude that the computational modeling shows that our data shows interference. Compared to the three-cues model proposed by \textcite{mertzen}, the Bayes factors provide moderate evidence for the new two-cues model. Therefore, the assumption that the relevant syntactic cue is the \{$\pm$subject-in-same-clause\} cue – rather than the previously assumed \{$\pm$grammatical subject\} cue – is moderately supported by these modeling results. This result holds for prior assumptions that allows higher probability mass for small values of $F$. i.e., for $F<0.01$.
